---
title: "Gene Expression Dataset Selection"
output:
  html_document:
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

# Preparation

```{r}
# Install the GEOmetadb package
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
    BiocManager::install("GEOmetadb")

getwd()
# DBI and knitr packages are also required to manipulate and visualize the query
```

Prepare the database file for the GEO database:

The directory where the database is located in the local disk. Currently
the database is used for the purpose of this course, so it is downloaded
into the R project directory of the container, 
i.e. (home/rstudio/projects/Jielin_Yang/).

```{r}
# Check if a database already exists
if(!file.exists("GEOmetadb.sqlite")) 
    GEOmetadb::getSQLiteFile()
# Display the version of the database
file.info("GEOmetadb.sqlite")

# Connect to the database using an absolute path to the file
con <- DBI::dbConnect(RSQLite::SQLite(), 
                      'GEOmetadb.sqlite')
```

# Query the GEOmetadb database

For database queries, a SQL query is constructed to select the
appropriate series from GEO.

In particular, we query the title, the technology used, the gse record
number, the date of deposit to GEO, as well as the links to the
supplementary files. We are looking for studies/series related to human
heart regeneration, therefore the organism must contain `Homo sapiens`,
and the title of the series must contain one of `heart`, `cardiac`,
`cardiomyocyte`, and one of `regeneration`, `proliferation`, and
`failure`.

```{r}
# Build the SQL query
sql <- paste("SELECT DISTINCT gse.title, gse.gse, gpl.title,",
             "gse.submission_date,  gse.supplementary_file",
             "FROM gse",
             "JOIN gse_gpl ON gse_gpl.gse=gse.gse",
             "JOIN gpl ON gse_gpl.gpl=gpl.gpl",
             "WHERE gse.submission_date > '2018-01-01'",
             "AND (gse.title LIKE '%heart%' OR gse.title LIKE '%cardiac%'",
             "OR gse.title LIKE '%cardiomyocyte%')",
             "AND (gse.title LIKE '%regenerat%' OR gse.title LIKE '%proliferat%'",
             "OR gse.title LIKE '%fail%')",
             "AND gpl.technology LIKE '%high-throughput%'",
             "AND gpl.organism LIKE '%Homo sapiens%'",
             sep=" ")

# Inidiate the query
result <- DBI::dbGetQuery(con, sql)
```

How many records are obtained?

```{r}
# Number of records
dim(result)[1]
```

Then we visualize the result by compiling a html table. Note, the links
to the supplementary files are obtained, but are not displayed in the
table. We will investigate the study/series first then check of the
study have the required files we need.

```{r}
# Visualize the retrieved series
knitr::kable(result[ , -5], table.attr = "class=\"striped\"", format = "html")
```
